A Framework for Discrete Change

### 2008-12-05 papers/individual/individual.tex
Updated individual paper, almost complete discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@246 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% phenomena of discrete change in development/learning
Piagetian developmental theory assumes step-wise changes in the rules
that children apply in all kinds of tasks such as in conservation
learning and in the balance scale task (SieglerAlibali19??).
\citet{Maas1992} developed a catastrophe model to describe step-wise
changes predicted to occur by Piagetian theory; importantly, they
derived precise statistical criteria for testing whether two
discretely different stages exist.  They applied the catastrophe model
to development on the conservation of liquid task (InhelderPiaget19??)
in which children have to judge relative volumes of liquid in glasses
of different heights and widths.  Young children tend to ignore the
width dimension and hence always choose the glass with the highest
level of liquid (InhelderPiaget19??).  It is assumed that there is a
sudden transition to a new rule, for example, a rule in which the
children also take the width of the glasses into account when judging
the volume of liquids.

% \citet{Maas1992} 
% opzoeken: Siegler & Alibali 19??, intro textbook
% opzoeken: Inhelder & Piaget, 19??

% balance scale rules
Jansen \& Maas (2001) applied the catastrophe model to
development of reasoning on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \citet{Siegler1981}.  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 \cite{Siegler1981}.

% Jansen 2001: aanpassen in de paper ref ipv het proefschrift

% hysteresis on the balance scale
\citet{Jansen2001} found evidence for stage-wise transitions, or
discrete change in our model, between Rule 1 and Rule 2 by testing
criteria that were derived from the catastrophe model.  In particular,
she found bimodal test scores and inaccessibility.  The latter means
that there are no in-between strategies: children apply either Rule 1
or Rule 2 and there is no in-between option.  \citet{Jansen2001} also
found evidence for hysteresis: the phenomenon that switching between
strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different values of a continuously
changing independent variable.  In particular, if the distance
dimension in the balance scale problems is made increasingly more
salient by increasing the distance difference between weights on
either side of the balance scale, children may switch from Rule 1 to
Rule 2.  If subsequently the distance difference is decreased again,
children may switch back to using Rule 1.  Hysteresis is the
phenomenon that this switch back occurs at a different value of the
control variable, in this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior.  For example, \citet{Gallistel2004}
found evidence for sudden onset of learning: at the start of the
learning experiment, pigeons did not learn anything and performance
was stable; after a number of trials, learning kicked in and there
were large increases in performance.  \citet{Gallistel2004}
focused on modeling the distribution of onset times: that is, the
trials at which learning suddenly takes off.  A similar interest in
process onset times is found in addiction research.  For example,
\citet{Sher2004} study the age at which children start using alcohol
and how this related to eventual outcomes in terms of addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  \citet{Raijmakers2001} found evidence for two
different strategies applied by children when faced with such a
learning task.  \citet{Schmittmann2006} reanalyzed their data using
hidden Markov models to show that both strategies are characterized by
sudden transitions in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, we give a
formalization of dependent mixture models and a brief overview of the
\pkg{depmixS4} package that was developed to specify and fit such
models.

@@

% outline
In this section we describe a class of models which are especially
suitable for describing and testing discrete change in (individual)
time series data.  The dependent mixture model is similar to, but
slightly different from, two other types of models that are in use for
modelling discrete change: the hidden and the latent Markov models.

% history of Markov models/LMM applications
Markov models have been used in the social sciences; for example, in
analyzing language learning \cite{Miller1952,Miller1963} and in the
analysis of paired associate learning \cite{Wickens1982}.  In these
models, the focus is on survey type data: a few repeated measurements
taken in a large sample.  \citeA{Langeheine1990} discuss latent Markov
models and their use in sociology and political science (see also
McCutcheon, 1987).  Latent Markov models have been used in studying
development of math skills \citep{Collins1992}, and in medical
applications \citep{Reboussin1998}; \citet{Kaplan2008} provides an
overview of such models, also called stage-sequential models in the
developmental psychology literature.

\nocite{McCutcheon1987}

@@

% depmix model
The dependent mixture model (DMM) that we propose here spans the range
from latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses
and covariates.  The dependent mixture model for a time series $O_{t}$
consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
\end{enumerate}
Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
\cite{Elliott1995}.

% state space
The state space, which is a set of discrete states, captures the
different states that the learning or developmental process under
consideration can be in.  In the balance scale example mentioned
above, children are applying one of two possible rules in responding
to the items.  The states are characterized by their corresponding
observation densities.  Using a particular rule in the balance scale
task leads to correct answers on some items and incorrect answers on
others.  A different rule may lead to correct answers on some items
and to guessing behavior on other items.  As another example, in
analyzing categorization learning data, in which participants learn to
categorize a set of objects, a typical initial state is that
participants are guessing because at the start of the task they have
no knowledge of which features are important in categorization.
Hence, the states in the state space represent knowledge states of the
participants, such that different knowledge states lead to different
observed behaviors or responses.

@@

% log-likelihood: kan dit helemaal weg???
The log-likelihood of DMMs is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \citet{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
Pr(\vc{O}_{t}|\vc{O}_{1},
	\ldots, \vc{O}_{t-1}),
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a
simple, i.e.\ observed, Markov chain these probabilities reduce to
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots,
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \citet{Lystig2002} define the following
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1})
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots,
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times
(\Phi_{t-1})^{-1},
	\label{eq:fwdt}
\end{split}
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and
equation~(\ref{eq:condLogl}) gives the following expression for the
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

Note that so far no assumptions have been made about the response
distributions $b_{j}$, hence these can be arbitrary univariate or
multivariate distributions.

@@

% outline/abstract: this has redundant information, mostly discussed
% in detail below
\pkg{depmixS4} implements a general framework for defining and fitting
dependent mixture models in the R programming language \citep{R2008}.
This includes standard Markov models, latent/hidden Markov models, and
latent class and finite mixture distribution models.  The models can
be fitted on mixed multivariate data with multinomial and/or gaussian
distributions.

@@

\pkg{depmixS4} is built using S4 classes (object oriented classes in R)
to allow easy extensibility (Chambers, 1998). A depmix model consists 
of the following: 
\begin{enumerate} 
	\item A model for the initial or prior probabilities 
	\item A list of transition models; one model for each row of the 
	transition matrix
	\item A list of response models for each of the states of the model; 
	this is a list of lists in the case of multivariate responses. 
\end{enumerate} 
Each of these are adressed briefly below.

@@

% response models
Response models in \pkg{depmixS4} interface with the \code{glm} 
functions available in R and with the \pkg{nnet}-package for the 
multinomial logistic models. Normally distributed data can hence be 
modelled with direct effects included as well. For example, normal 
data are modelled as:
\begin{equation}
	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
\end{equation}
where $\mu$ is the (state-dependent) mean; $\vc{x}$ is a vector of
covariates or predictors; $\mathbf{\beta}$ is the vector of
(state-dependent) regression coefficients; $\epsilon_{t}$ is a
normally distributed error term with (state-dependent) standard
deviation $\sigma$.

@@

% MAARTEN: de multinomial logistic models maken weer gebruik van andere
% packages toch, kun je daar de referentie van geven?

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  EM is used by
default in unconstrained models, but otherwise direct optimization is
done using \pkg{Rdonlp2} \cite{Tamura2007,Spellucci2002}, because it
handles general linear (in)equality constraints, and optionally also
non-linear constraints.

@@

Two illustrations are provided below of models that analyze single
participant time series data from two common experimental paradigms.
In both of these, participants learn different strategies through
trial and error.

@@

% what is the HDT? developmental trends and relevance?
\citet{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the absolute amounts were reduced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples).  At doors A and C, a loss of 10 apples (A) or 2 apples (C)
is delivered in 50\% of the trials.  For doors B and D, frequency of
loss is only 10\%.  The median loss of doors B and D is 50 and 10,
respectively.  \citet{Crone2004} administered the HDT to children from
four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
that children have difficulty to take future consequences into account.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007b} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most of the youngest children seem
to focus frequency of loss, resulting in equal preference for doors B
and D. It's assumed that frequency of loss is the dominant dimension
in the task.  Older participants seem to use a two-dimensional rule
where participants first focus on the frequency of loss and then
consider amount of loss, resulting in a preference for door D. A third
very small subgroup seems to use an integrative rule where
participants combine all three dimensions in the appropriate way.
Participants using the integrative rule pick cards from doors C and D,
which are advantageous in the long run.

@@

% problematic aspects of standard analysis
\citet{Huizenga2007b} used the last 60 trials in a series of 200
trials, assuming that behavior has stabilized after 140 trials of
learning; this could very well be wrong and it is highly likely that
there are individual differences in this learning process.  Hence,
here we analyze IGT learning data from a single participant with the
aim to establish 1) at which trial learning stops, i.e., at which
trial behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
various participants.  The responses were fitted with multinomial
logistic models for the four possible choices that participants make in
this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that there were
designated begin and end states.  This means that the initial state
probabilities were fixed to one for the first state and to zero for
the remaining states.  For the transition matrix this means that the
final state was an absorbing state: the probability of transitioning
out of that state is zero.  This was done to ensure that there would
be a final state in which participants end which provides the
possibility of immediately seeing participants' behavior at the end of
the task.  Models were selected using the Akaike Information Criterion
(AIC; Akaike, 1973).

\nocite{Akaike1973} % check the year!!

The data were best described by a 4-state model.  This model's
transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4} below.

@@

\begin{table}
	\caption{Estimates for the Iowa Gambling Task model}
	\label{tab:igt4}
	\begin{tabular}{lcccccccc} \hline
		state & p(A) & p(B) & p(C) & p(D) \\
		$S_1$ & 0    & 0.06 & 0.31 & 0.63 \\
		$S_2$ & 0.33 & 0.67 & 0    & 0 \\ 
		$S_3$ & 0    & 0    & 1    & 0 \\
		$S_4$ & 0    & 0.06 & 0.94 & 0 
	\end{tabular}
\end{table}

The states have clear interpretations.  States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses.  States 1 and 2 are dominated by C and D, and by A and B
responses, respectively.  To get a clearer interpretation of these
states it is necessary to consider when they are visited during the
learning process.  The Viterbi algorithm \citep{Rabiner1989} provides
us with the posterior state sequence, i.e., the sequence of states
that the participant is in at each trial.  This sequence is depicted
in Figure~\ref{fig:post4}.

@@

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{graphs/post4.pdf}
	\label{fig:post4}
	\caption{Posterior state sequence for the 4 state model.}
\end{center}
\end{figure}

At the start of training there are many switches between states 1 and 
2, indicating that behavior is mostly random choices between all four 
categories, albeit with a slight preference for B and D over A and C
choices. This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D both have low frequency of
loss. After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3 
behavior, consisting only of C responses, a short transitional period 
and then state 4 behavior consisting of 94 \% C choices and some B
choices, i.e., almost optimal behavior. The choice proportions and the
model predicted choice proportions are depicted in
Figure~\ref{fig:igt4}.

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{graphs/igt4.pdf}
	\label{fig:igt4}
	\caption{Data and model predicted proportions of door choices in
	the IGT.}
\end{center}
\end{figure}

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices represents one of the optimal strategies, which consists of
both C and D responses. Note that both C and D responses generate equal
profits in the long run.

@@

The Weather Prediction Task \cite<WPT>{Knowlton1994} %(WPT, Knowlton, Squire \& Gluck, 1994) 
is a probabilistic categorization task in which participants learn to
predict (or categorize) the state of the weather (sun or rain) on the
basis of four ``tarot'' cards (cards with abstract geometrical
patterns).  On a given trial, one, two, or three cues are present.
There are a total of 14 possible cue patterns, and each cue pattern is
associated with a particular probability distribution over the states
of the weather.  In order to perform well in the task, participants
must predict the weather in accordance with these conditional
probabilities.

>>>>

% phenomena of discrete change in development/learning
Piagetian developmental theory assumes step-wise changes in the rules
that children apply in all kinds of tasks such as in conservation
learning and in the balance scale task \citep{Siegler2005}.
\citet{Maas1992} developed a catastrophe model to describe step-wise
changes predicted to occur by Piagetian theory; importantly, they
derived precise statistical criteria for testing whether two
discretely different stages exist.  They applied the catastrophe model
to development on the conservation of liquid task \citep{Inhelder1958}
in which children have to judge relative volumes of liquid in glasses
of different heights and widths.  Young children tend to ignore the
width dimension and hence always choose the glass with the highest
level of liquid \citep{Inhelder1958}.  It is assumed that there is a
sudden transition, or a discrete change, to a new rule, for example, a
rule in which the children also take the width of the glasses into
account when judging the volume of liquids.

% balance scale rules
\citet{Jansen2001b} applied the catastrophe model to development of
reasoning on the balance scale task \citep{Siegler1981}.  In the
balance scale task participants have to judge which side of a balance
goes down when the number of weights and their distances to the
fulcrum are varied over trials.  Younger children tend to ignore the
distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \citep{Siegler1981}.
Older children include the distance dimension in determining their
response to balance scale problems; however, they only do so when the
weight dimension does not differ between the sides of the balance,
i.e., when the number of weights is equal on both sides of the balance
scale.  This strategy is called Rule 2 \cite{Siegler1981}.

% hysteresis on the balance scale
\citet{Jansen2001b} found evidence for stage-wise transitions, or
discrete change in our model, between Rule 1 and Rule 2 by testing
criteria that were derived from the catastrophe model.  In particular,
they found bimodal test scores and inaccessibility.  The latter means
that there are no in-between strategies: children apply {\em either}
Rule 1 {\em or} Rule 2 and there is no in-between option.
\citet{Jansen2001b} also found evidence for hysteresis: the phenomenon
that switching between strategies is asymmetric.  Children can switch
from Rule 1 to Rule 2 and back, but this occurs at different values of
a continuously changing independent variable.  In particular, if the
distance dimension in the balance scale problems is made increasingly
more salient by increasing the distance difference between weights on
either side of the balance scale, children may switch from Rule 1 to
Rule 2.  If subsequently the distance difference is decreased again,
children may switch back to using Rule 1.  Hysteresis is the
phenomenon that this switch back occurs at a different value of the
control variable, in this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for
discrete changes in response behavior.  For example,
\citet{Gallistel2004} found evidence for sudden onset of learning: at
the start of the learning experiment, pigeons did not learn anything
and performance was stable; after a number of trials, learning kicked
in and there were large increases in performance.
\citet{Gallistel2004} focused on modeling the distribution of onset
times: that is, the trials at which learning suddenly takes off.  A
similar interest in process onset times is found in addiction
research.  For example, \citet{Sher2004} study the age at which
children start using alcohol and how this relates to eventual outcomes
in terms of (severity of) addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  \citet{Raijmakers2001} found evidence for two
different strategies applied by children when faced with such a
learning task.  \citet{Schmittmann2006} reanalyzed their data using
hidden Markov models to show that both strategies are characterized by
discrete changes in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009c} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
Weather Prediction Task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, we give a
formalization of dependent mixture models and a brief overview of the
\pkg{depmixS4} package that was developed to specify and fit such
models.

@@

% outline
In this section we describe a class of models which are especially
suitable for describing and testing discrete changes in (individual)
time series data.  The dependent mixture model is similar to, but
slightly different from, two other types of models that are in use for
modelling discrete change: the hidden and the latent Markov models.

% history of Markov models/LMM applications
Markov models have been used for a long time in the social sciences;
for example, in analyzing language learning
\cite{Miller1952,Miller1963} and in the analysis of paired associate
learning \cite{Wickens1982}.  In these models, the focus is on survey
type data: a few repeated measurements taken in a large sample.
\citeA{Langeheine1990} discuss latent Markov models and their use in
sociology and political science \citep<see also>{McCutcheon1987}.  Latent
Markov models have been used in studying development of math skills
\citep{Collins1992}, and in medical applications
\citep{Reboussin1998}.  \citet{Kaplan2008} provides an overview of
such models, also called stage-sequential models, and their
application in developmental psychology.

@@

% depmix model
The dependent mixture model (DMM) that we propose here spans the range
from latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses
and covariates.  The dependent mixture model for a time series $O_{t}$
consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states, forming the state space.
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix.
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
\end{enumerate}
Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
\cite{Elliott1995}.

% state space
The state space $S$ captures the different states that the learning or
developmental process under consideration can be in.  In the balance
scale example mentioned above, children are applying one of two
possible rules in responding to the items.  The states are
characterized by their corresponding observation densities.  Using a
particular rule in the balance scale task leads to correct answers on
some items and incorrect answers on others.  A different rule may lead
to correct answers on some items and to guessing behavior on others.
As another example, in analyzing categorization learning data, in
which participants learn to categorize a set of objects, a typical
initial state is that participants are guessing because at the start
of the task they have no knowledge of which features are important in
categorization.  Hence, the states in the state space represent
knowledge states of the participants, such that different knowledge
states lead to different observed behaviors or responses.

@@

% % log-likelihood: kan dit helemaal weg???
% The log-likelihood of DMMs is usually computed by the
% so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
% rather by the forward part of this algorithm.  \citet{Lystig2002}
% changed the forward algorithm in such a way as to allow computing the
% gradients of the log-likelihood at the same time. 
% 
% 
% They start by rewriting the likelihood as follows (for ease of
% exposition the dependence on the model parameters is dropped here):
% \begin{equation}
% 	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
% Pr(\vc{O}_{t}|\vc{O}_{1},
% 	\ldots, \vc{O}_{t-1}),
% 	\label{condLike}
% \end{equation}
% where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a
% simple, i.e.\ observed, Markov chain these probabilities reduce to
% $Pr(\vc{O}_{t}|\vc{O}_{1},\ldots,
% \vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
% The log-likelihood can now be expressed as:
% \begin{equation}
% 	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
% \vc{O}_{t-1})].
% 	\label{eq:condLogl}
% \end{equation}
% 
% To compute the log-likelihood, \citet{Lystig2002} define the following
% (forward) recursion:
% \begin{align}
% 	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1})
% 	\label{eq:fwd1} \\
% \begin{split}
% 	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots,
% \vc{O}_{t-1}) \\
% 	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times
% (\Phi_{t-1})^{-1},
% 	\label{eq:fwdt}
% \end{split}
% \end{align}
% where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining
% $\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and
% equation~(\ref{eq:condLogl}) gives the following expression for the
% log-likelihood:
% \begin{equation}
% 	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
% 	\label{eq:logl}
% \end{equation}
% 
% Note that so far no assumptions have been made about the response
% distributions $b_{j}$, hence these can be arbitrary univariate or
% multivariate distributions.

@@

% outline/abstract: this has redundant information, mostly discussed
% in detail below
\pkg{depmixS4} \citep{Visser2008a} implements a general framework for
defining and fitting dependent mixture models in the R programming
language \citep{R2008}.  This includes standard Markov models,
latent/hidden Markov models, and latent class and finite mixture
distribution models.  The models can be fitted on mixed multivariate
data with error distriutions from the generalized linear model
framework, such as the gaussian, binomial, multinomial logistic et
cetera.

@@

\pkg{depmixS4} is built using S4 classes (object oriented classes in R)
to allow easy extensibility (Chambers, 1998). A depmix model consists 
of the following: 
\begin{enumerate} 
	\item A model for the initial or prior probabilities.
	\item A list of transition models; one model for each row of the 
	transition matrix.
	\item A list of response models for each of the states of the model; 
	this is a list of lists in the case of multivariate responses. 
\end{enumerate} 
Each of these are adressed briefly below.

@@

% response models
Response models in \pkg{depmixS4} interface with the \code{glm}
functions \citep{R2008} available in R and with the \pkg{nnet}-package
\citep{Venables2002} for the multinomial logistic models.  Normally
distributed data can hence be modelled with direct effects included as
well.  For example, normal data are modelled as:
\begin{equation}
	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
\end{equation}
where $\mu$ is the (state-dependent) mean; $\vc{x}$ is a vector of
covariates or predictors; $\mathbf{\beta}$ is the vector of
(state-dependent) regression coefficients; $\epsilon_{t}$ is a
normally distributed error term with (state-dependent) standard
deviation $\sigma$.

@@

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
alternatively through the use of a general Newton-Raphson optimizer.
EM is used by default in unconstrained models, but otherwise direct
optimization is done using \pkg{Rdonlp2}
\cite{Tamura2007,Spellucci2002}, because it handles general linear
(in)equality constraints, and optionally also non-linear constraints.
Both the forward algorithm \citep{Baum1966,Rabiner1989} and the
version by \citet{Lystig2002} are implemented in \pkg{depmixS4}; the
latter is implemented to allow easy computation of the gradients of
the parameters.

@@

Two illustrations are provided below of models that analyze single
participant time series data from two common experimental paradigms.
In both of these, participants learn different strategies through
trial and error.  The main goal of these illustrations is to establish
the possibility of analyzing single participant data using the
\pkg{depmixS4}-framework.

@@

% what is the HDT? developmental trends and relevance?
\citet{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the absolute amounts were reduced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples) at each trial.  At doors A and C, a loss of 10 apples (A) or 2
apples (C) is delivered in 50\% of the trials.  For doors B and D,
frequency of loss is only 10\%.  The median loss of doors B and D is
50 and 10, respectively.  \citet{Crone2004} administered the HDT to
children from four age groups (6-9, 10-12, 13-15, and 18-25 year-olds)
and concluded that children have difficulty to take future
consequences into account.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007b} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most of the youngest children seem
to focus on frequency of loss, resulting in equal preference for doors
B and D. It is assumed that frequency of loss is the dominant
dimension in this task.  Older participants seem to use a
two-dimensional rule where participants first focus on the frequency
of loss and then consider amount of loss, resulting in a preference
for door D. A third very small subgroup seems to use an integrative
rule where participants combine all three dimensions in the
appropriate way.  Participants using the integrative rule pick cards
from doors C and D, which are both equally advantageous in the long
run, although door D shows a higher volatility.

@@

% problematic aspects of standard analysis
\citet{Huizenga2007b} used the last 60 trials in a series of 200
trials, as is typically done in analyzing this task.  Doing so
silently assumes that behavior has stabilized after 140 trials of
learning; this may not be the case, and it is highly likely that there
are individual differences in this learning process.  Hence, here we
analyze IGT learning data from a single participant with the aim to
establish 1) at which trial learning stops, i.e., at which trial
behavior has stabilized, and 2) what strategies participants use
during and at the end of learning.  Models with an increasing number
(from 2 through 5) of latent states were fitted to the time series of
a single participant.  The responses were fitted with multinomial
logistic models for the four possible choices that participants make
in this task.  There were no covariates on any of the parameters.  The
only constraint imposed on the models' parameters was that there were
designated begin and end states.  This means that the initial state
probabilities were fixed to one for the first state and to zero for
the remaining states.  For the transition matrix this means that the
final state was an absorbing state: the probability of transitioning
out of the final state was set to zero.  This was done to ensure that
there would be a final state in which participants end which provides
the possibility of immediately seeing participants' behavior at the
end of the task.  Models were selected using the Akaike Information
Criterion \citep<AIC;>{Akaike1973}.

The data were best described by a 4-state model.  This model's
transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix, the final state is
absorbing.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~\ref{tab:igt4} below.

@@

\begin{table}
	\caption{Estimates for response probabilities in the IGT model.}
	\label{tab:igt4}
	\begin{tabular}{lcccccccc} \hline
		state & p(A) & p(B) & p(C) & p(D) \\
		$S_1$ & 0    & 0.06 & 0.31 & 0.63 \\
		$S_2$ & 0.33 & 0.67 & 0    & 0 \\ 
		$S_3$ & 0    & 0    & 1    & 0 \\
		$S_4$ & 0    & 0.06 & 0.94 & 0 
	\end{tabular}
\end{table}

The states have clear interpretations.  States 3 and 4 are both
dominated by C responses, and state 4 has a low probability of B
responses.  States 1 and 2 are dominated by C and D, and by A and B
responses, respectively.  To get a clearer interpretation of these
states, it is necessary to consider when they are visited during the
learning process.  The Viterbi algorithm \citep{Rabiner1989} provides
us with the posterior state sequence, i.e., the sequence of states
that the participant is in at each trial.  This sequence is depicted
in Figure~1.

@@

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{graphs/post4.pdf}
	\caption{Posterior state sequence for the 4 state model of the IGT 
	data.}
	\label{fig:post4}
\end{center}
\end{figure}

At the start of training there are many switches between states 1 and
2, indicating that behavior is mostly random choices between all four
categories, albeit with a slight preference for B and D over A and C
choices.  This preference indicates a focus on the frequency of loss
associated with each of the doors as B and D both have low frequency
of loss.  After this, there is a period of stable state 2 behavior
associated with responses A and B. After that there is stable state 3
behavior, consisting only of C responses, a short transitional period
and then state 4 behavior consisting of 94 \% C choices and some B
choices, i.e., almost optimal behavior.  The choice proportions and
the model predicted choice proportions are depicted in Figure~2.

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{graphs/igt4.pdf}
	\caption{Data and model predicted proportions of door choices in
	the IGT.}
	\label{fig:igt4}
\end{center}
\end{figure}

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007b} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices represents one of the optimal strategies, which consists of
both C and D responses. Note that C and D responses generate equal
profits in the long run.

@@

The Weather Prediction Task \cite<WPT;>{Knowlton1994} is a
probabilistic categorization task in which participants learn to
predict (or categorize) the state of the weather (sun or rain) on the
basis of four ``tarot'' cards (cards with abstract geometrical
patterns).  On a given trial, one, two, or three cues are present.
There are a total of 14 possible cue patterns, and each cue pattern is
associated with a particular probability distribution over the states
of the weather.  In order to perform well in the task, participants
must predict the weather in accordance with these conditional
probabilities.
